Simulating High-Dimensional Multivariate Data

### 2021-02-26 040-monte-carlo-evaluation.Rmd
renv+box fully implemented

```{r ch040-LoadLib040, include=FALSE}
devtools::load_all()
library(tidyverse)
## library(bigsimr)
## Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
## bs <- bigsimr_setup()
## dist <- distributions_setup()
```

@@

```{r ch040-biGammaPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
bivariate_gamma_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
## ggsave('fig/plot-biGammaPlot.pdf')
```

@@

## AGS 12/22/20: Removing the latex formatting fixes the html render.
## But how to improve latex table formatting? Better to use the template
tabMAE %>%
    knitr::kable(booktabs = TRUE,
                 col.names = c("No. of random vectors",
                               "Correlation type",
                               "Distribution",
                               "Mean abs. error"),
                 caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.')
```

@@

To mimick the workflow, we first produce a synthetic "data set" by completing the following steps:  
1. Produce heterogeneous gamma marginals by randomly selecting the $j^{th}$ gamma shape parameter from $U_j \sim uniform(1,10), j=1,\ldots,d$ and the $j^{th}$ rate parameter from $V_j \sim exp(1/5), j=1,\ldots,d$, with the constant parameters determined arbitrarily.  
2. Produce a random full-rank Pearson correlation matrix via `cor_randPD` of size $d \times d$.  
3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec` (without matching the Pearson exactly).

With the synthetic data set in hand, we complete and time the following 4 steps:  
i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.  
1. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.  
2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.  
3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.

@@

tabTimes %>%
    knitr::kable(booktabs = TRUE,
                 col.names = c("Dimension",
                               "Correlation type",
                               "Total Time (Seconds)" ),
                 caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
```

The results larger dimensional vectors show scalebility to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
Figure \@ref(ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$. For $d$ of 1000 and 2500, the total time is under a couple of minutes. At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly. Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*. Uniformly, matching Spearman correlations is faster with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type. With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtaned the 1,000 vectors in under an hour (data not shown). In principle, this would enable the simulation of an entire human-derived RNA-seq data set.
We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, *Simulate Data*, is fast under all schemes for all $d$ under consideration.

```{r ch040-largeDfig, echo=F, fig.cap="Computation times as d increases."}
# Set colors
## RColorBrewer::display.brewer.all()
numColors <- 5
numGroups <- 4
myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])

>>>>

```{r ch040-LoadLib040, include=FALSE}
box::use(
    dplyr[...],
    ggplot2[...],
    patchwork[...],
    tidyr[pivot_longer],
    bigsimr[bigsimr_setup, distributions_setup],
    knitr[kable],
    RColorBrewer[brewer.pal]
)
load("data/bivariate_normal_sims.rda")
load("data/bivariate_gamma_sims.rda")
load("data/bivariate_nbinom_sims.rda")
load("data/benchmark_dependences.rda")

Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
bs <- bigsimr_setup(pkg_check = FALSE)
dist <- distributions_setup()
prob <- 3e-04
```

@@

```{r ch040-biGammaPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
bivariate_gamma_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
```

@@

kable(tabMAE, booktabs = TRUE,
      col.names = c("No. of random vectors",
                    "Correlation type",
                    "Distribution",
                    "Mean abs. error"),
      caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
```

@@

To mimick the workflow, we first produce a synthetic "data set" by completing the following steps:

1. Produce heterogeneous gamma marginals by randomly selecting the $j^{th}$ gamma shape parameter from $U_j \sim uniform(1,10), j=1,\ldots,d$ and the $j^{th}$ rate parameter from $V_j \sim exp(1/5), j=1,\ldots,d$, with the constant parameters determined arbitrarily.  
2. Produce a random full-rank Pearson correlation matrix via `cor_randPD` of size $d \times d$.  
3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec` (without matching the Pearson exactly).

With the synthetic data set in hand, we complete and time the following 4 steps:

i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.  
1. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.  
2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.  
3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.

@@

tabTimes %>%
    kable(booktabs = TRUE,
          col.names = c("Dimension",
                        "Correlation type",
                        "Total Time (Seconds)"),
          caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
```

The results larger dimensional vectors show scalebility to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$. For $d$ of 1000 and 2500, the total time is under a couple of minutes. At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly. Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*. Uniformly, matching Spearman correlations is faster with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type. With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtaned the 1,000 vectors in under an hour (data not shown). In principle, this would enable the simulation of an entire human-derived RNA-seq data set. We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, *Simulate Data*, is fast under all schemes for all $d$ under consideration.

```{r ch040-largeDfig, fig.cap="Computation times as d increases."}
# Set colors
numColors <- 5
numGroups <- 4
myColors <- rev(brewer.pal(n=numColors, name="Greys")[((numColors-numGroups)+1):numColors])

### 2021-02-26 050-examples.Rmd
renv+box fully implemented

```{r ch050-LoadLib, include=FALSE, cache=FALSE}
box::use(
  dplyr[...],
  ggplot2[...],
  patchwork[...],
  bigsimr[...],
  knitr[include_graphics]
)
load("data/example_brca.rda")
load("data/example_genes.rda")

Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
bs <- bigsimr_setup(pkg_check = FALSE)
dist <- distributions_setup()

set.seed(10142020)
```

@@

```{r ch050-nbHelpers, echo=TRUE}
make_nbinom_margins <- function(sizes, probs) {
  margins <- lapply(1:length(sizes), function(i) {
    dist$NegativeBinomial(sizes[i], probs[i])
  })
  do.call(c, margins)
}

nbinom_mom <- function(x) {
  m <- mean(x)
  s <- sd(x)
  s2 <- s^2
  c(size = m^2 / (s2 - m), prob = m/s2)
}
```

@@

```{r ch050-estMargins, echo = TRUE, eval = TRUE, cache=F}
nbinom_fit <- apply(brca1000, 2, nbinom_mom)
sizes <- nbinom_fit["size",]
probs <- nbinom_fit["prob",]
nb_margins <- make_nbinom_margins(sizes, probs)
```

@@

# 2a. Check admissibility
(is_valid_corr <- JuliaCall::julia_call("Bigsimr.iscorrelation", R_X))

@@

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache=FALSE}
sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
colnames(sim_nbinom) <- colnames(brca1000)
```

```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache=FALSE}
dir.create("results", showWarnings = FALSE, recursive = TRUE)
brca_sim_path <- paste0("results/brca_", d, "_sim.rds")
if(!file.exists(brca_sim_path)) {
  set.seed(44)
  sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
  colnames(sim_nbinom) <- colnames(brca1000)
  saveRDS(sim_nbinom, brca_sim_path)
}
sim_nbinom <- readRDS(brca_sim_path)
```

@@

```{r ch050-simDataFig, cache=FALSE, fig.cap="Simulated data for three selected high-expressing genes generally replicates the estimated data structure. The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values."}
set.seed(02242021)
num_genes    <- 3
gene_sample  <- sample(example_genes[1:1000], num_genes)

GGally::ggpairs(
  data = as.data.frame(sim_nbinom[, gene_sample]),
  upper = list(continuous = GGally::wrap('cor', method = "spearman"))
) + theme_bw()
```

@@

<!-- 
Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.
your comment -->

```{r ch050-makeFigBRCA, cache=FALSE}
R_S_hat <- bs$cor(as.matrix(sim_nbinom), bs$Spearman)
sim_nbinom_fit <- apply(sim_nbinom, 2, nbinom_mom)
corr_compare <- tibble(
  `Target Correlation` = R_S[lower.tri(R_S, diag = FALSE)],
  `Estimated Correlation` = R_S_hat[lower.tri(R_S_hat, diag = FALSE)]
)
target_compare <- tibble(
  `Target Size` = sizes,
  `Estimated Size` = sim_nbinom_fit[1,]
)
logprob_compare <- tibble(
  `Target Log Probability` = log(probs),
  `Estimated Log Probability` = log(sim_nbinom_fit[2,])
)

if(!file.exists("fig/ch050-figBRCA.png")) {
  p1 <- ggplot(corr_compare, aes(`Target Correlation`, `Estimated Correlation`)) +
    geom_point(alpha = 0.05) + 
    geom_abline(slope = 1, col = "red", lty = "dashed") +
    theme_bw() + coord_equal()

  p2 <- ggplot(target_compare, aes(`Target Size`, `Estimated Size`)) +
    geom_point(alpha = 0.85) + 
    geom_abline(slope = 1, col = "red", lty = "dashed") +
    theme_bw() + coord_equal()

  p3 <- ggplot(logprob_compare, 
               aes(`Target Log Probability`, `Estimated Log Probability`)) +
    geom_point(alpha = 0.85) + 
    geom_abline(slope = 1, col = "red", lty = "dashed") +
    theme_bw() + coord_equal()

  p <- p1 + p2 + p3

  ggsave(filename = "fig/ch050-figBRCA.png",
         plot = p,
         device = "png",
         width = 8,
         height = 4,
         units = "in")
}
```

@@

```{r ch050-densityEvaluation, include=FALSE, cache=FALSE}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
y0 <- 1
pHat <- mean(apply(sim_nbinom, 1, function(Y) {
  all(Y >= y0)
}))
```

@@

```{r ch050-densityEvaluationECHO, echo=TRUE, cache=FALSE}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
y0 <- 1
mean(apply(sim_nbinom, 1, function(Y) {
  all(Y >= y0)
}))
```

@@

```{r ch010-myM, include=FALSE}
m <- 10
```

@@

In this small example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$. Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-gpuVScpuFig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite). At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{Spearman}$, producing a distribution of loss values.

```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE, cache=FALSE}
# Simulate random vectors equal to the sample size
N <- nrow(example_brca)
# create m random vectors and estimate correlation
simRho <- replicate(n = m, expr = {
  tmpSim <- bs$rvec(N , R_X, nb_margins)
  bs$cor(tmpSim, bs$Spearman)
}, simplify = FALSE)
# Evaluate the residual sum of squared error
sapply(simRho, function(R) sum((R - R_S)^2))
```

```{r loss_examples, include=FALSE, eval=FALSE}
sapply(simRho, function(R) norm(R - R_S))           # norm of diff
sapply(simRho, function(R) norm(R) - norm(R_S))     # diff of norms
sapply(simRho, function(R) sum((R - R_S)^2))        # RSS
sapply(simRho, function(R) mean((R - R_S)^2))       # MSE
sapply(simRho, function(R) sqrt(mean((R - R_S)^2))) # RMSE
sapply(simRho, function(R) mean(abs(R - R_S)))      # MAE
```

```{r ch050-lossResults, cache=FALSE}
frobenius_path <- paste0("results/brca_", d, "_frobenius.rds")
if (!file.exists(frobenius_path)) {
    N <- nrow(example_brca)
    ## create m random vectors and estimate correlation
    simRho <- replicate(n = m, expr = {
        tmpSim <- bs$rvec(N, R_X, nb_margins)
        bs$cor(tmpSim, bs$Spearman)
    }, simplify = FALSE)
    frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
    saveRDS(frobenius_loss, frobenius_path)
}
frobenius_loss <- readRDS(frobenius_path)
```

@@

```{r ch050-loss-summary, echo=TRUE, eval=TRUE, cache=FALSE}
summary(frobenius_loss)
```

>>>>

```{r ch050-LoadLib, include=FALSE, cache=FALSE}
box::use(
  dplyr[...],
  ggplot2[...],
  GGally[ggpairs, wrap],
  patchwork[...],
  bigsimr[...],
  knitr[include_graphics],
  JuliaCall[julia_call],
  ./R/utils[mom_nbinom]
)
load("data/example_brca.rda")
load("data/example_genes.rda")
sim_nbinom <- readRDS("results/brca_1000_sim.rds")
simRho <- readRDS("results/brca_1000_frobenius.rds")

Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
bs <- bigsimr_setup(pkg_check = FALSE)
dist <- distributions_setup()
```

@@

```{r ch050-nbHelpers, echo=TRUE}
make_nbinom_margins <- function(sizes, probs) {
  margins <- lapply(1:length(sizes), function(i) {
    dist$NegativeBinomial(sizes[i], probs[i])
  })
  do.call(c, margins)
}
```

@@

```{r ch050-estMargins, echo=TRUE}
nbinom_fit <- apply(brca1000, 2, mom_nbinom)
sizes <- nbinom_fit["size",]
probs <- nbinom_fit["prob",]
nb_margins <- make_nbinom_margins(sizes, probs)
```

@@

# 2a. Check admissibility
(is_valid_corr <- julia_call("Bigsimr.iscorrelation", R_X))

@@

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE}
sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
colnames(sim_nbinom) <- colnames(brca1000)
```

@@

```{r ch050-simDataFig, cache=FALSE, fig.cap="Simulated data for three selected high-expressing genes generally replicates the estimated data structure. The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values."}
set.seed(2020-02-25)
num_genes <- 3
gene_sample <- sample(example_genes[1:1000], num_genes)

ggpairs(data = as.data.frame(sim_nbinom[, gene_sample]),
        upper = list(continuous = wrap('cor', method = "spearman"))) + 
  theme_bw()
```

@@

<!-- 
Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.
-->

@@

```{r ch050-densityEvaluation, include=FALSE}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
y0 <- 1
pHat <- mean(apply(sim_nbinom, 1, function(Y) {
  all(Y >= y0)
}))
```

@@

```{r ch050-densityEvaluationECHO, echo=TRUE}
<<ch050-densityEvaluation>>
pHat
```

@@

@@

In this small example, we simulate $m=30$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$. Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-largeDfig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite). At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{Spearman}$, producing a distribution of loss values.

```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE}
# Simulate random vectors equal to the sample size
N <- nrow(example_brca)
# create m random vectors and estimate correlation
simRho <- replicate(n = m, expr = {
  tmpSim <- bs$rvec(N , R_X, nb_margins)
  bs$cor(tmpSim, bs$Spearman)
}, simplify = FALSE)
# Evaluate the residual sum of squared error
sapply(simRho, function(R) sum((R - R_S)^2))
```

```{r ch050-lossResults}
frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
```

@@

```{r ch050-loss-summary, echo=TRUE}
summary(frobenius_loss)
```
